### 2020-07-20 code/figure1.Rmd
modify plotting to coord_fixed for umaps; added plots 1b and 1d, rename umap coordinates UMAP 1 and UMAP 2

cluster_umap <- umap(mca_res$ind$coord,
  min_dist = 1e-6,
  n_components = 3,
  n_neighbors = round(sqrt(nrow(mca_res$ind$coord))),
  n_epochs = 1000,
  n_threads = 1
)
colnames(cluster_umap) <- c("cluster_x", "cluster_y", "cluster_z")
hdb_cluster_assignment <- hdbscan(cluster_umap, minPts = optimal_minPts)
umap_3d_coors <- gdc_umap_coors %>%
  bind_cols(as.data.frame(cluster_umap)) %>%
  mutate(
    clust = as.factor(hdb_cluster_assignment$cluster),
    member_prob = hdb_cluster_assignment$membership_prob,
    sample_id = rownames(gdc_umap_coors)
  )
## clust 0 is unclassified cluster

@@

permut_p <- randomize %>%
  filter(eigenvalue < 50) %>%
  ggplot(aes(x = eigenvalue, y = expl_var / 100)) +
  geom_line(aes(colour = "#998ec3")) +
  geom_line(aes(y = perm / 100, colour = "#f1a340")) +
  scale_color_identity(guide = guide_legend(title = ""), 
                       labels = c("Eigenvalues", "Permuted Eigenvalues")) +
  scale_y_continuous(labels = scales::percent) +
  xlab("MCA Eigenvalues") +
  ylab("explained variance") +
  theme_classic()

pvalue_p <- randomize %>%
  filter(eigenvalue < 50) %>%
  ggplot(aes(x = eigenvalue, y = pvalue)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = optimal_dimensions, color = "red") +
  ggtitle(label = "", subtitle = paste0("Significant dimensions = ", optimal_dimensions)) +
  xlab("MCA Eigenvalues") +
  ylab("P-value") +
  theme_classic()

@@

## umap monochrome 2d
fig1C <- ggplot(umap_cancertype, aes(x=plot_x, y=plot_y)) +
  geom_point(colour = "grey", size = 0.25) +
  xlab("umap_1") +
  ylab("umap_2") +
  theme_classic() +
  labs(tag = "C") +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank())
## umap hdb_scan
fig1E <- ggplot(subset(umap_cancertype, clust !=0), aes(x = plot_x, y = plot_y, color = as.factor(clust))) +
  geom_point(data = umap_cancertype, aes(x = plot_x, y = plot_y), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) + scale_color_manual(values = kelly.colours[-c(1:2)]) +
  guides(color = F) +
  theme_classic() +
  xlab("umap_1") +
  ylab("umap_2") +
  theme_classic() +
  labs(tag = "E") +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank())

## umap knn_clust 
fig1F <- ggplot(subset(umap_cancertype), aes(x = plot_x, y = plot_y, color = as.factor(knn_clust))) +
  geom_point(size = 0.25) + scale_color_manual(values = kelly.colours[-c(1:2)]) +
  guides(color = F) +
  theme_classic() +
  xlab("umap_1") +
  ylab("umap_2") +
  theme_classic() +
  labs(tag = "F") +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank())

@@

## umap by cancer type
fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(x = plot_x, y = plot_y, colour = project_code)) +
  geom_point(data = umap_cancertype, aes(x = plot_x, y = plot_y), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) + 
  theme_classic() +
  ylab("umap_2") +
  xlab("umap_1") +
  labs(tag = "H", color = "") +
  scale_color_manual(labels = c("BRCA", "COAD", "HNSC", "PAAD"), 
                     values = c('#fb9a99', '#b15928', '#6a3d9a', 'springgreen4')) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank(),
        legend.position = "bottom")

## umap of brca subtype
fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(x = plot_x, y = plot_y, colour = Subtype_Selected)) +
  geom_point(data = umap_cancertype, aes(x = plot_x, y = plot_y), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) +
  theme_classic() +
  ylab("umap_2") +
  xlab("umap_1") +
  labs(tag = "I", color = "") +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank(),
        legend.position = "bottom")

@@

fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(x = plot_x, y = plot_y, colour = mut_type)) +
  geom_point(data = brca, aes(x = plot_x, y = plot_y), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) + 
  theme_classic() +
  ylab("umap_2") +
  xlab("umap_1") +
  labs(color = "", tag = "J") +
  scale_color_manual(labels = c("BRCA1", "BRCA1,BRCA2", "BRCA2"), 
                     values = c('#33a02c', '#e31a1c', '#ff7f00')) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank(),
        legend.position = "bottom")

## umap stage
fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(x = plot_x, y = plot_y, colour = ajcc_pathologic_tumor_stage)) +
  geom_point(data = umap_cancertype, aes(x = plot_x, y = plot_y), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) + 
  theme_classic() +
  ylab("umap_2") +
  xlab("umap_1") +
  labs(tag = "K", color = "") +
  scale_color_manual(labels = c("Stage I", "Stage II", "Stage III", "Stage IV"), 
                     values = c('darkgoldenrod3','dodgerblue3','yellow3','tomato2')) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank(),
        legend.position = "bottom")

## umap metastasis
fig1L <- ggplot(umap_cancertype, aes(x = plot_x, y = plot_y)) +
  geom_point(data = umap_cancertype, aes(x = plot_x, y = plot_y), colour = "grey", size = 0.25) +
  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(x = plot_x, y = plot_y, colour = ajcc_metastasis_pathologic_pm), size = 0.25) + 
  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(x = plot_x, y = plot_y, colour = ajcc_metastasis_pathologic_pm), size = 0.25) + 
  ylab("umap_2") +
  xlab("umap_1") +
  labs(tag = "L", color = "") +
  scale_color_manual(labels = c("M0", "M1"), 
                     values = c('#a6cee3','#1f78b4')) +
  theme_classic() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank(),
        legend.position = "bottom")

layout <- '
AAAABBCCDD
AAAAEEEFFF
AAAAGGGGGG
AAAAGGGGGG
AAAAHHIIJJ
AAAAKKLL##
'

fig1A + plot_spacer() + fig1C + plot_spacer() + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout)

>>>>

cluster_umap <- umap(mca_res$ind$coord,
  min_dist = 1e-6,
  n_components = 3,
  n_neighbors = round(sqrt(nrow(mca_res$ind$coord))),
  n_epochs = 1000,
  n_threads = 1
)
colnames(cluster_umap) <- c("cluster_x", "cluster_y", "cluster_z")
hdb_cluster_assignment <- hdbscan(cluster_umap, minPts = optimal_minPts)
umap_3d_coors <- gdc_umap_coors %>%
  bind_cols(as.data.frame(cluster_umap)) %>%
  mutate(
    n_pathways_mut = rowSums(gdc_path[, -1]),
    clust = as.factor(hdb_cluster_assignment$cluster),
    member_prob = hdb_cluster_assignment$membership_prob,
    sample_id = rownames(gdc_umap_coors)
  )
## clust 0 is unclassified cluster

@@

load("../data/figure1B_data.rda")
optimal_mca_dimensions <- randomize %>%
  filter(pvalue < 0.05) %>%
  filter(eigenvalue == max(eigenvalue)) %>%
  pull(eigenvalue)

permut_p <- randomize %>%
  filter(eigenvalue < 50) %>%
  ggplot(aes(x = eigenvalue, y = expl_var / 100)) +
  geom_line(aes(colour = "#998ec3")) +
  geom_line(aes(y = perm / 100, colour = "#f1a340")) +
  scale_color_identity(guide = guide_legend(title = ""), 
                       labels = c("Eigenvalues", "Permuted Eigenvalues")) +
  scale_y_continuous(labels = scales::percent) +
  xlab("MCA Eigenvalues") +
  ylab("explained variance") +
  theme_classic() +
  theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12))

pvalue_p <- randomize %>%
  filter(eigenvalue < 50) %>%
  ggplot(aes(x = eigenvalue, y = pvalue)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = optimal_dimensions, color = "red") +
  ggtitle(label = "", subtitle = paste0("Significant dimensions = ", optimal_dimensions)) +
  xlab("MCA Eigenvalues") +
  ylab("P-value") +
  theme_classic() +
  theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12))

@@

fig1C <- ggplot(umap_cancertype %>% arrange(n_pathways_mut), 
                aes(x=plot_x, y=plot_y)) +
  geom_point(size = 0.25) +
  xlab("UMAP 1") +
  ylab("UMAP 2") +
  coord_fixed() +
  theme_classic() +
  scale_color_viridis_c()
  labs(tag = "C") +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank())

load("../data/optimal_minpts_score.rda")
fig1D <- ggplot(score_df, aes(x = minPts, y = score)) +
  geom_point(data = sample_df, color = "grey", aes(y = value)) +
  geom_point() + 
  geom_line() +
  geom_vline(xintercept = optimal_minPts, color = "#ff0000") +
  scale_y_log10() +
  xlab("Minimum Cluster Size") +
  ggtitle(paste0("Optimal Minimum Cluster Size = ", optimal_minPts)) +
  theme_classic() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank())

## umap hdb_scan
fig1E <- ggplot(subset(umap_cancertype, clust !=0), aes(x = plot_x, y = plot_y, color = as.factor(clust))) +
  geom_point(data = umap_cancertype, aes(x = plot_x, y = plot_y), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) + scale_color_manual(values = kelly.colours[-c(1:2)]) +
  guides(color = F) +
  theme_classic() +
  xlab("UMAP 1") +
  ylab("UMAP 2") +
  coord_fixed() +
  theme_classic() +
  labs(tag = "E") +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank())

## umap knn_clust 
fig1F <- ggplot(subset(umap_cancertype), aes(x = plot_x, y = plot_y, color = as.factor(knn_clust))) +
  geom_point(size = 0.25) + scale_color_manual(values = kelly.colours[-c(1:2)]) +
  guides(color = F) +
  theme_classic() +
  xlab("UMAP 1") +
  ylab("UMAP 2") +
  coord_fixed() +
  theme_classic() +
  labs(tag = "F") +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank())

@@

## umap by cancer type
fig1H <- ggplot(umap_cancertype[umap_cancertype$project_code %in% c("COAD", "BRCA", "HNSC", "PAAD"),], aes(x = plot_x, y = plot_y, colour = project_code)) +
  geom_point(data = umap_cancertype, aes(x = plot_x, y = plot_y), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) + 
  theme_classic() +
  ylab("UMAP 2") +
  xlab("UMAP 1") +
  coord_fixed() +
  labs(tag = "H", color = "") +
  scale_color_manual(labels = c("BRCA", "COAD", "HNSC", "PAAD"), 
                     values = c('#fb9a99', '#b15928', '#6a3d9a', 'springgreen4')) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank(),
        legend.position = "bottom")

## umap of brca subtype
fig1I <- ggplot(umap_cancertype %>% filter(project_code == "BRCA" & !is.na(Subtype_Selected)), aes(x = plot_x, y = plot_y, colour = Subtype_Selected)) +
  geom_point(data = umap_cancertype, aes(x = plot_x, y = plot_y), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) +
  theme_classic() +
  ylab("UMAP 2") +
  xlab("UMAP 1") +
  coord_fixed() +
  labs(tag = "I", color = "") +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank(),
        legend.position = "bottom")

@@

fig1J <- ggplot(brca %>% filter(mut_type != "NA"), aes(x = plot_x, y = plot_y, colour = mut_type)) +
  geom_point(data = brca, aes(x = plot_x, y = plot_y), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) + 
  theme_classic() +
  ylab("UMAP 2") +
  xlab("UMAP 1") +
  coord_fixed() +
  labs(color = "", tag = "J") +
  scale_color_manual(labels = c("BRCA1", "BRCA1,BRCA2", "BRCA2"), 
                     values = c('#33a02c', '#e31a1c', '#ff7f00')) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank(),
        legend.position = "bottom")

## umap stage
fig1K <- ggplot(subset(umap_cancertype, ajcc_pathologic_tumor_stage != "Not Available"), aes(x = plot_x, y = plot_y, colour = ajcc_pathologic_tumor_stage)) +
  geom_point(data = umap_cancertype, aes(x = plot_x, y = plot_y), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) + 
  theme_classic() +
  ylab("UMAP 2") +
  xlab("UMAP 1") +
  coord_fixed() +
  labs(tag = "K", color = "") +
  scale_color_manual(labels = c("Stage I", "Stage II", "Stage III", "Stage IV"), 
                     values = c('darkgoldenrod3','dodgerblue3','yellow3','tomato2')) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank(),
        legend.position = "bottom")

## umap metastasis
fig1L <- ggplot(umap_cancertype, aes(x = plot_x, y = plot_y)) +
  geom_point(data = umap_cancertype, aes(x = plot_x, y = plot_y), colour = "grey", size = 0.25) +
  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M0"), aes(x = plot_x, y = plot_y, colour = ajcc_metastasis_pathologic_pm), size = 0.25) + 
  geom_point(data = subset(umap_cancertype, ajcc_metastasis_pathologic_pm == "M1"), aes(x = plot_x, y = plot_y, colour = ajcc_metastasis_pathologic_pm), size = 0.25) + 
  ylab("UMAP 2") +
  xlab("UMAP 1") +
  coord_fixed() +
  labs(tag = "L", color = "") +
  scale_color_manual(labels = c("M0", "M1"), 
                     values = c('#a6cee3','#1f78b4')) +
  theme_classic() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.ticks = element_blank(),
        legend.position = "bottom")

layout <- '
AAAABBCCDD
AAAABBEEFF
AAAAGGGGGG
AAAAGGGGGG
AAAAHHIIJJ
AAAAKKLL##
'

fig1A + fig1B + fig1C + fig1D + fig1E + fig1F + fig1G + fig1H + fig1I + fig1J + fig1K + fig1L + plot_layout(design = layout)
